---
title: "Appendix"
author: "Peter Hickey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: ../latex/header.tex
  html_document:
    keep_md: true
bibliography: ../latex/phd_thesis.bib
---

## The probability that two dependent Bernoulli random variables are identical

\cite{Lindqvist:1978fm} wrote a brief note on Bernoulli trials with dependence. 
Building on earlier work by \cite{Klotz:1973if}, Lindqvist parameterises the 
Bernoulli process $X_{1}, X_{2}, \ldots$ on $\{0, 1\}$ by the parameters 
$p = Pr(X_{i} = 1)$ and $c = cor(X_{i - 1}, X_{i})$ and shows that the 
transition matrix is given by

\begin{equation*}
\label{eq:transition_matrix_id}
\Pi =
 \begin{pmatrix}
  (1 - p) + cp & p (1 - c) \\
  (1 - p) (1 - c) & p + c (1 - p)
 \end{pmatrix}
\end{equation*}

provided that $max(1 - \frac{1}{p}, 1 - \frac{1}{1 - p}) \leq c \leq 1$. 

From this we can compute the joint distribution,

\begin{equation*}
Pr(X_{1} = x_{1}, \ldots, X_{n} = x_{n}) = Pr(X_{1} = x_{1}) Pr(X_{2} = x_{2} | X_{2} = x_{2}) \cdots Pr(X_{n} = x_{n} | X_{n - 1} = x_{n - 1})
\end{equation*}

In particular, in the case $n = 2$ we can compute the probability that two 
dependent but identically distributed Bernoulli random variables are equal. 

To extend the above result to the probability that two dependent and 
__non-identically__ distributed Bernoulli random variables are equal, simply 
requires that we derive the appropriate transition matrix. Switching notation 
to that used in Chapter \ref{chap:co-methylation_review}, let 
$Z_{h, i} \eqd Bernoulli(p_{h})$ and $Z_{h', i'} \eqd Bernoulli(p_{h'})$. The 
transition matrix, 
$\Pi = \bigl(\begin{smallmatrix} 
Pr(Z_{h', i'} = z_{h', i'} | Z_{h, i} = z_{h, i}) 
\end{smallmatrix} \bigr)$, is given by

\begin{equation*}
\label{eq:transition_matrix_nid}
\Pi =
 \begin{pmatrix}
  (1 - p_{h'}) + c p_{h} & p_{h'} - c p_{h} \\
  \big [ (1 - p_{h'}) + c p_{h} \big ] (1 - p_{h}) & \big [ p_{h'} + c (1 - p_{h}) \big ] p_{h}
 \end{pmatrix}
\end{equation*}

We can then compute the desired probability

\begin{align*}
Pr(Z_{h, i} = Z_{h', i'}) &= Pr(Z_{h, i} = 0, Z_{h', i'} = 0) + Pr(Z_{h, i} = 1, Z_{h', i'} = 1) \\
                  &= Pr(Z_{h', i'} = 0 | Z_{h, i} = 0) Pr(X_{0} = 0) +
                     Pr(Z_{h', i'} = 1 | Z_{h, i} = 1) Pr(X_{0} = 1) \\
                  &= \big [ (1 - p_{h'}) + c p_{h} \big ] (1 - p_{h}) +
                     \big [ p_{h'} + c ( 1 - p_{h}) \big ] p_{h} \\
                  &= (1 - p_{h}) (1 - p_{h'}) + c p_{h} \big [1 - p_{h}] +
                     p_{h} p_{h'} + c p_{h} (1 - p_{h}) \\
                  &= (1 - p_{h}) (1 - p_{h'}) + p_{h} p_{h'} + 2 c p_{h} (1 - p_{h})
\end{align*}

__TODO: What is the general restriction on $c$ when $p_{h, i} \neq p_{h', i'}$?__

## Analysis of \cite{Lacey:2013iy}

I want to estimate the number of CpGs used by \cite{Lacey:2013iy} to estimate the correlation of methylation levels. They used the chromosome 11 data from an RRBS experiment on a normal myotube cell line, MTCTL2. Unfortunately, the data are not publicly available. Instead, I use data from ENCODE of a skeletal muscle myotubes cell line that had RRBS performed in triplicate (HSMMtube).

Firstly, I download the data and unzip it:
```{r, cache = FALSE, message = FALSE}
library(R.utils)
library(data.table)
library(dplyr)
print(getwd())
```

```{r, cache = FALSE, message = FALSE}
dl_dir <- 'downloads/'
url <- 'http://hgdownload.cse.ucsc.edu/goldenPath/hg19/encodeDCC/wgEncodeHaibMethylRrbs/'
files <- paste0('wgEncodeHaibMethylRrbsHsmmtubefshdDukeSitesRep',
                1:3, '.bed.gz')
dests <- paste0(dl_dir, files)
x <- mapply(function(f, d) {
  if (!file.exists(gsub(".gz", "", paste0(dl_dir, f)))) {
    download.file(url = paste0(url, f), d)
    gunzip(filename = d, overwrite = TRUE)
    }
  fread(gsub(pattern = '.gz', replacement = '', x = d), sep = '\t', 
        verbose = FALSE)
  }, files, dests, SIMPLIFY = FALSE)
```

Then I count the number of CpGs on chromosome 11 in each sample:
```{r}
y <- lapply(x, function(xx) {
  xx <- tbl_dt(xx)
  xx %>%
    group_by(V1) %>%
    summarise(n())
})

rbind_all(y) %>% filter(V1 == 'chr11')
```


### TODOs

* Cite data source and publication
