---
title: "Appendix"
author: "Peter Hickey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: ../latex/header.tex
  html_document:
    keep_md: true
bibliography: ../latex/phd_thesis.bib
---

## The probability that two dependent Bernoulli random variables are identical

\cite{Lindqvist:1978fm} wrote a brief note on Bernoulli trials with dependence. Building on earlier work by \cite{Klotz:1973if}, Lindqvist parameterises the Bernoulli process $X_{1}, X_{2}, \ldots$ on $\{0, 1\}$ by the parameters $p = Pr(X_{i} = 1)$ and $c = cor(X_{i - 1}, X_{i})$ and shows that the transition matrix is given by

\begin{equation*}
\label{eq:transition_matrix_id}
\Pi =
 \begin{pmatrix}
  (1 - p) + cp & p (1 - c) \\
  (1 - p) (1 - c) & p + c (1 - p)
 \end{pmatrix}
\end{equation*}

From this we can compute the joint distribution,

\begin{equation*}
Pr(X_{1} = x_{1}, \ldots, X_{n} = x_{n}) = Pr(X_{1} = x_{1}) Pr(X_{2} = x_{2} | X_{2} = x_{2}) \cdots Pr(X_{n} = x_{n} | X_{n - 1} = x_{n - 1})
\end{equation*}

In particular, in the case $n = 2$ we can compute the probability that two dependent but identically distributed Bernoulli random variables are equal, $Pr(X_{1} = X_{2}) = Pr(X_{1} = 0, X_{2} = 0) + Pr(X_{1} = 1, X_{2} = 1)$.

To extend the above result to the probability that two dependent but __non-identically__ distributed Bernoulli random variables are equal simply requires that we derive the appropriate transition matrix.

Let $X_{i} \eqd Bernoulli(p_{i})$ for $i = 1, 2$. The transition matrix, $\Pi = \bigl(\begin{smallmatrix}
Pr(X_{2} = x_{2} | X_{1} = x_{1}) \end{smallmatrix} \bigr)$, is given by

\begin{equation*}
\label{eq:transition_matrix_nid}
\Pi =
 \begin{pmatrix}
  (1 - p_{2}) + c p_{1} & p{2} - c p_{1} \\
  \big [ (1 - p_{2}) + c p_{1} \big ] (1 - p_{1}) & \big [ p_{2} + c (1 - p_{1}) \big ] p_{1}
 \end{pmatrix}
\end{equation*}

We can then compute the desired probability

\begin{align*}
Pr(X_{1} = X_{2}) &= Pr(X_{1} = 0, X_{2} = 0) + Pr(X_{1} = 1, X_{2} = 1) \\
                  &= Pr(X_{2} = 0 | X_{1} = 0) Pr(X_{0} = 0) +
                     Pr(X_{2} = 1 | X_{1} = 1) Pr(X_{0} = 1) \\
                  &= \big [ (1 - p_{2}) + c p_{1} \big ] (1 - p_{1}) + 
                     \big [ p_{2} + c ( 1 - p_{1}) \big ] p_{1} \\
                  &= (1 - p_{1}) (1 - p_{2}) + c p_{1} \big [1 - p_{1}] +
                     p_{1} p_{2} + c p_{1} (1 - p_{1}) \\
                  &= (1 - p_{1}) (1 - p_{2}) + p_{1} p_{2} + 2 c p_{1} (1 - p_{1})
\end{align*}

__TODO: What is the general restriction on $c$ when $p_{h, i} \neq p_{h', i'}$?__

## Analysis of \cite{Lacey:2013iy}

I want to estimate the number of CpGs used by \cite{Lacey:2013iy} to estimate the correlation of methylation levels. They used the chromosome 11 data from an RRBS experiment on a normal myotube cell line, MTCTL2. Unfortunately, the data are not publicly available. Instead, I use data from ENCODE of a skeletal muscle myotubes cell line that had RRBS performed in triplicate (HSMMtube).

Firstly, I download the data and unzip it:
```{r, cache = TRUE, message = FALSE}
library(R.utils)
library(data.table)
library(dplyr)

tmp <- tempdir()
url <- 'http://hgdownload.cse.ucsc.edu/goldenPath/hg19/encodeDCC/wgEncodeHaibMethylRrbs/'
files <- paste0(url, 'wgEncodeHaibMethylRrbsHsmmtubefshdDukeSitesRep', 
                     1:3, '.bed.gz')
dests <- paste0(tmp,'/rep', 1:3, '.bed.gz')
x <- mapply(function(f, d) {
  download.file(f, d)
  gunzip(filename = d, overwrite = TRUE)
  fread(gsub(pattern = '.gz', replacement = '', x = d), sep = '\t')
  }, files, dests, SIMPLIFY = FALSE)
```

Then I count the number of CpGs on chromosome 11 in each sample:
```{r}
y <- lapply(x, function(xx) {
  xx <- tbl_dt(xx)
  xx %>% 
    group_by(V1) %>% 
    summarise(n())
})

rbind_all(y) %>% filter(V1 == 'chr11')
```


### TODOs
 
* Cite data source and publication

